PEACE: {\underline P}arallel {\underline E}nvironment for {\underline A}ssembly and {\underline C}lustering of Gene {\underline E}xpression

### 2010-02-14 papers/NAR_web_server_2009/paper.tex
Added a missing image that I had missed out earlier.

\begin{abstract}
  We present \peace, a stand alone tool for the high-throughput {\it
    ab initio} clustering of transcript fragment sequences produced by
  Next Generation or Sanger Sequencing technologies, freely available
  from \href{http://www.peace-tools.org}{www.peace-tools.org} .
  Installed and run through an easily downloaded GUI, \peace\/ can
  process large data sets of transcript fragments of length 50 bases
  or greater, grouping the fragments by gene association and achieving
  a greater sensitivity (without sacrificing runtime) than the
  competing clustering tools.  Once clustered, the user can use the GUI
  to collect statistics and single out specific clusters for more
  comprehensive study.  \peace\/, when used as a ``pre-assembly''
  tool, enables any assembly tool to work with individual small
  clusters of closely related sequences.  The result is a significant
  increase in sensitivity, improved robustness to sequencing error,
  and a reduction in runtime.  For a 5\% sequencing error rate, we see
  a 52\% improvement in sensitivity over the leading cluster tool,
  \wcd\/, and 184\% improvement in sensitivity over the assembly tool
  \capthree.  When applied to a 361 Mb set of Arabidopsis transcriptome
  fragments, we see a 40\% speedup over \wcd.
\end{abstract}

@@

\begin{figure*}
  \begin{minipage}{3in}
    \includegraphics[width=3in]{screen.d/install_page_big.png}
    \centerline{\small{(a)}}
  \end{minipage}
  \begin{minipage}{2in}
    \vspace{0.47in}
    \includegraphics[width=2in]{screen.d/add_server_big.png}
    \centerline{\small{(b)}}
  \end{minipage}
  \begin{minipage}{2in}
    \vspace{0.47in}
    \includegraphics[width=2in]{screen.d/add_server_big.png}
    \centerline{\small{(c)}}
  \end{minipage}

@@

\peace\/ has been tested on both simulated and real data from NGS and
Sanger Sequencing technologies, comparing results against those
produced by the \wcd\/ clustering tool \cite{Hazelhurst08a} and the
\capthree\/ assembly tool \cite{Huang99} (the latter of which
implicitly calculates a clustering in the process of assembly).  For our
simulation tests we used the {\bf ESTSim} tool \cite{Hazelhurst03} to
generate simulated transcript fragments of varying length under
different models of error (Supplementary Materials, Section~C.1),
generating the fragments from the list of 100 zebra fish genes used in
the \wcd\/ testing \cite{Hazelhurst08a}.  Tool parameters were taken
to match, as closely as possible, those used in the \wcd\/ study (see
supplementary materials).  Our primary methods of quality assessment
were {\it sensitivity} (the fraction of fragment pairs from the same
gene that were correctly clustered together) and {\it Type 1 error}
(the fraction of genes that were divided between clusters)
\cite{Wang04,Hazelhurst08a}.  Specificity was not an issue in
simulated tests, as both \peace\/ and \wcd\/ were completely successful
in separating unrelated ESTs (save when dealing with recently
duplicated genes).  Other measurements are also discussed in the
supplementary materials.  In Figure~\ref{SeT1} (left), reflecting
tests done on simulated Sanger EST sets, we see a
significant improvement of \peace\/ (blue) over \wcd\/ (green) and
\capthree\/ (black) in its sensitivity to sequencing errors, while in
Figure~\ref{SeT1} (right) we see a comparable improvement in Type 1
error.  In Figure~\ref{singletons} we look at the number of singleton
clusters (fragments not joined to any cluster), which should not occur
in our simulated sets.  We again see significant improvement in \peace.
The sequential runtime of \peace\/ is slightly improved over that of
\wcd\/ (see Supplementary Materials, Figure~S4).

@@

A comparative test of \peace\/ against \wcd\/ on NGS short-read sequences
was not feasible due to the implementation of \wcd.  In theory, that
tool should be able to handle short-read sequences with the
appropriate parameter settings.  However, in the distributed versions those
parameters seemed to be hard-coded, thus rendering testing impossible.
\peace\/ was run on both simulated and real NGS fragment sets.  For the
former we found \peace\/ to do well, achieving a sensitivity of 0.75
when simulating a sequencing error rate of 2\% (with comparable {\it
  Jaccard Index} value, which reflects both sensitivity {\it and}
specificity \cite{Kalyanaraman03,Hazelhurst08a}).  Runtime was
approximately 17.6 minutes to process the 7.9 Mb of data when run in
sequential mode on a 3.2 GHz Intel Xeon CPU.

>>>>

\begin{abstract}
  We present \peace, a stand alone tool for the high-throughput {\it
    ab initio} clustering of transcript fragment sequences produced by
  Next Generation or Sanger Sequencing technologies, freely available
  from \href{http://www.peace-tools.org}{www.peace-tools.org}.
  Installed and run through an easily downloaded GUI, \peace\/ can
  process large data sets of transcript fragments of length 50 bases
  or greater, grouping the fragments by gene association and achieving
  a greater sensitivity (without sacrificing runtime) than the
  competing clustering tools.  Once clustered, the user can employ the
  GUI to collect statistics and single out specific clusters for more
  comprehensive study.  \peace\/, when used as a ``pre-assembly''
  tool, enables any assembly tool to work with individual small
  clusters of closely related sequences.  For a 5\% sequencing error
  rate, we see a 52\% improvement in sensitivity over the leading
  cluster tool, \wcd\/, and 184\% improvement in sensitivity over the
  assembly tool \capthree.  When applied to a 361 Mb set of
  Arabidopsis transcriptome fragments, we see a 40\% speedup over
  \wcd.
\end{abstract}

@@

\begin{figure*}
  \begin{minipage}{3in}
    \includegraphics[width=3in]{screen.d/install_page_big.png}
    \centerline{\small{(a)}}
  \end{minipage}
  \begin{minipage}{2in}
    \vspace{0.47in}
    \includegraphics[width=2in]{screen.d/add_server_big.png}
    \centerline{\small{(b)}}
  \end{minipage}
  \begin{minipage}{2in}
    \vspace{0.66in}
    \includegraphics[width=2in]{screen.d/processor_assignment_big.png}
    \centerline{\small{(c)}}
  \end{minipage}

@@

\peace\/ has been tested on both simulated and real data from NGS and
Sanger Sequencing technologies, comparing results against those
produced by the \wcd\/ clustering tool \cite{Hazelhurst08a} and the
\capthree\/ assembly tool \cite{Huang99} (the latter of which
implicitly calculates a clustering in the process of assembly).  For our
simulation tests we used the {\bf ESTSim} tool \cite{Hazelhurst03} to
generate simulated transcript fragments of varying length under
different models of error (Supplementary Materials, Section~C.1),
generating the fragments from the list of 100 zebra fish genes used in
the \wcd\/ testing \cite{Hazelhurst08a}.  Tool parameters were taken
to match, as closely as possible, those used in the \wcd\/ study (see
Supplementary Materials).  Our primary methods of quality assessment
were {\it sensitivity} (the fraction of fragment pairs from the same
gene that were correctly clustered together) and {\it Type 1 error}
(the fraction of genes that were divided between clusters)
\cite{Wang04,Hazelhurst08a}.  Specificity was not an issue in
simulated tests, as both \peace\/ and \wcd\/ were completely successful
in separating unrelated ESTs (save when dealing with recently
duplicated genes).  Other measurements are also discussed in the
supplementary materials.  In Figure~\ref{SeT1} (left), reflecting
tests done on simulated Sanger EST sets, we see a
significant improvement of \peace\/ (blue) over \wcd\/ (green) and
\capthree\/ (black) in its sensitivity to sequencing errors, while in
Figure~\ref{SeT1} (right) we see a comparable improvement in Type 1
error.  In Figure~\ref{singletons} we look at the number of singleton
clusters (fragments not joined to any cluster), which should not occur
in our simulated sets.  We again see significant improvement in \peace.
The sequential runtime of \peace\/ is slightly improved over that of
\wcd\/ (see Supplementary Materials, Figure~S4).

@@

We were unable to run \wcd\/ on short-read sequences, as the
distributed implementation is unable to handle such data without an
adjustment to the code.  Hence, while \peace\/ was run on both
simulated and real NGS fragment sets, there was no cluster tool to
which the results could be compared.  \peace\/ did well, achieving a
sensitivity of 0.75 when applied to a simulated short-read set subjected to
a sequencing error rate of 2\% (with comparable {\it Jaccard Index}
values reflecting both sensitivity {\it and} specificity
\cite{Kalyanaraman03,Hazelhurst08a}).  Runtime was approximately 17.6
minutes to process the 7.9 Mb of data when run in sequential mode on a
3.2 GHz Intel Xeon CPU.  When applied to a 64 Mb set of 165,490 {\it
  Chlamydomonas reinhardtii} transcript fragments sequenced by 454
pyrosequencing technology (ranging in size from 50 bases to several
hundred), \peace\/ completed the job in 49 minutes using 30 of the
Xeon processors.
